Solving Consensus with OpenFlow

### 2015-02-26 inband-STN.tex
bla

\title{One-Shot Coordination of Distributed SDN Controllers:\\On the Consensus Power of OpenFlow}

@@

\begin{abstract}
The control plane of future Software-Defined Networks (SDNs)
is likely to be \emph{distributed}: a redundant control plane
can ensure a high availability, and a spatially distributed
control plane allows to handle time-critical dataplane
events close to their origin. Moreover, future SDNs may also
allow multiple administrators or participatory users to update
policies concurrently, e.g., by offering transactional interfaces. 
However, not much is known today about how to implement 
a consistent distributed control plane which supports concurrent
policy installations.

In this paper, we propose to coordinate distributed controllers
directly on the Openflow switches: exploiting classic results
on multi-writer consensus, we show how to use Openflow 
to synchronize controllers and install policies in a consistent
manner, \emph{in one shot}. Our approach is attractive as it
does not require complicated helping mechanisms in case of controller
failures. 
\end{abstract}

@@

While the logically centralized network perspective has the potential to simplify
the network management---many network tasks are inherently non-local---and introduces
many flexibilities,
managing and operating an SDN in a more adaptive but consistent manner is non-trivial, even with SDN.
A particularly interesting problem regards the consistent installation of new policies
or routes.~\cite{todo,todo,todo} While implementing consistent network updates 
is challenging already from the perspective of a single controller, there exists a wide consensus
that the logically centralized SDN control planes will likely be \emph{actually distributed} in the future.

The reason for implementing SDN control planes in a distributed manner are manyfold.
First, in order to provide a high availability as well as a minimal degree of
fault-tolerance, controllers should be redundant~\cite{onix,stn,todo}: a failure
of one controller can be masked by other controllers. Second, it has also been proposed
to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication intensive control plane events close to their origin.~\cite{devoflow,kandoo,jukka,disco}
Third, larger SDNs are likely to be operated by multiple administrators or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently.~\cite{participatory,stn}

However, today, we do not have a good understanding of how to realize
such distributed control planes. The problem is essentially a distributed
computing one: Multiple controllers may simultaneously try to
install conflicting policies, and a consensus must be reached which one of
the two is actually installed. Moreover, controllers may also fail,
and a mechanism must be provided to complete an update in this scenario.

@@

In this paper, we propose an approach to solve these problems.
In particular, we promote a very direct approach to coordinate controllers,
using the Openflow bundle feature to make an atomic consensus and update:
using one message, a single controller will be able to install its policy,
\emph{in one shot}. Our approach relies on a classic result in 
distributed computing, namely multi-writer consensus. Interestingly,
this result is not well-known even in the distributed computing community,
and we believe that the SDN paradigm offers a very practical
application for it. We believe that our approach nicely complements
existing, more theoretical literature on distributed control planes; for example,
it shows how to realize STN~\cite{stn}.

@@

ideas: distributed control planes

ideas: known consensus results and impossibilities

ideas: the power of in-band

@@

ideas: the missing link for STN, but also others

>>>>

%\title{One-Shot Coordination of Distributed SDN Controllers:\\On the Consensus Power of OpenFlow}

\title{Coordinating Distributed SDN Controllers:\\On the Consensus Power of OpenFlow}

@@

\begin{abstract}
The control plane of modern Software-Defined Networks (SDNs)
is \emph{distributed}: a redundant control plane
can ensure a high availability, and a spatially distributed
control plane allows to handle time-critical dataplane
events close to their origins. Moreover, an SDN may 
be managed by multiple administrators simultaneously. 
However, not much is known today about how to implement
such a distributed control plane. 
In this paper, we show that powerful coordination objects
can be implemented using today's Openflow switches.
In particular, we describe the implementation of test\&set primitives 
as well as multi-writer consensus objects which can be used
to implement atomic transactions with all-or-nothing semantics.
Interestingly, our approach allows controllers to detect and
resolve very general notions of conflicts, not only, e.g., conflicts
due to overlapping flow spaces but also conflicts based on network load.
\end{abstract}

@@

While the logically centralized network perspective has the potential to simplify
the network management---many network tasks are inherently non-local---and introduces
many flexibilities,
managing and operating an SDN in a more adaptive but consistent manner is non-trivial, even with SDN.
A particularly interesting problem regards the consistent installation of new policies
or routes.~\cite{network-update,roger-hotnets,correct} While implementing consistent network updates
is challenging already from the perspective of a single controller, there exists a wide consensus
that the logically centralized SDN control planes will likely be \emph{actually distributed} in the future.

The reason for implementing SDN control planes in a distributed manner are manyfold.
First, in order to provide a high availability as well as a minimal degree of
fault-tolerance, controllers should be redundant~\cite{onix,stn,onos}: a failure
of one controller can be masked by other controllers. Second, it has also been proposed
to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication intensive control plane events close to their origin.~\cite{devoflow,kandoo,jukka,disco}
Third, larger SDNs are likely to be operated by multiple administrators or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently.~\cite{participatory,stn}

However, today, we do not have a good understanding of how to realize
such distributed control planes. The problem is essentially a distributed
systems one: Multiple controllers may simultaneously try to
install conflicting updates, e.g., rules (of the same priority) defined over non-disjoint
flow space, or load-balancing policies which, if combined, actually \emph{increase} the load~\cite{log-cent}. 
Moreover, controllers may also fail,
before their update has been completed.

@@

In this paper, we show that powerful synchronization primitives
can be implemented in Openflow, allowing controllers to implement
non-trivial atomic transactions, and detect and
automatically resolve conflicts. In particular, we show how to implement
the classic distributed computing primitive
\emph{test\&set}, supporting the conditional installation of new rules,
and we show how to use the bundle feature to implement
multi-writer objects, an efficient means to compute consensus;
interestingly, our implementation can also be used to compute a consensus
and install the winner's rule \emph{atomically}, using one message.
Moreover, our approach allows to detect and resolve very general notions of conflicts:
conflicts may be defined, for example, for rules of the same priority over
defined over non-independent flow spaces, but may also depend, e.g., on 
network load (based on counters).
We believe that our approach nicely complements
existing, more theoretical literature on distributed control planes; for example,
it also shows how to realize the atomic read-modify-write primitive postulated in
STN~\cite{stn}.

@@

There exists a wide consensus that SDN control planes need to be distributed.~\cite{onos,onix,elasticon} 
Onix~\cite{onix} is one of the earliest proposals, and introduced the 
the notion of Network Information Base (NIB), abstracting network state
distribution from control logic, but 
requiring mechanisms for the detection and resolution of conflicts. 
Also spatially distributed control planes to improve scalability and
latency have been studied intensively
in the literature~\cite{kandoo,ctrl-place,hotsdn13loc}
proposes an elastic distributed controller architecture.

Operating and updating SDNs consistently is already non-trivial
from the perspective of a single controller. Reitblatt et al.~~\cite{network-update} 
introduced the notion of
per-packet consistency and proposed a 2-phase update protocol based on tagging.
%, and described the \emph{two-phase update} technique, also used in our algorithms.
Mahajan and Wattenhofer~\cite{roger-hotnets} considered weaker transient
consistency guarantees, and proposed more efficient network update algorithms
accordingly. Ludwig et al.~~\cite{hotnets14update} studied algorithms for secure
network updates where packets are forced to traverse certain waypoints or
middleboxes. Ghorbani et al.~~\cite{correct-virt} recently argued for the design
of network update algorithms that provide even stronger consistency guarantees.

To the best of our knowledge, the only paper explicitly addressing the consistent
network update problem from a distributed controller perspective is STN~\cite{stn}:
STN provides a transactional interface offering all-or-nothing semantics and serializability
(the ``holy grail'' of safety properties), 
allowing
controllers to install policies in a conflict-free manner; if some controllers fail,
other controllers can take over. STN~\cite{stn} is based on 
atomic
read-modify-write primitives, but while this primitive is postulated, no implementation is
described. In this paper, we provide this missing link, and also show that other useful and
much more powerful synchronization primitives can be implemented.

Our work is also closely related to the recent work on providing more high-level
abstractions and programming languages for SDNs, such as Frenetic~\cite{frenetic}, 
also enabling policy composition~\cite{pyretic}.

In terms of distributed computing: TODO PETR: maybe modify a bit the text below
and also write more about multi-writer consensus etc.?

\noindent\textbf{Distributed Computing.}
There is a long tradition of defining correctness of a concurrent system via
an equivalence to a sequential one~\cite{Pap79-serial,Lam79,HW90}.  The notion
of sequentially composable histories is reminiscent of
linearizability~\cite{HW90}, where a history of operations concurrently
applied by a collection of processes is equivalent to a history in which the
operations are in a sequential order, respecting their real-time precedence.
In contrast, our sequentially composable histories impose requirements not
only on high-level invocations and responses, but also on the way the traffic
is processed. We require that the committed policies constitute a
conflict-free sequential history, but, additionally,  we expect that each
\emph{path} witnesses only a prefix of this history, consisting of all
requests that were committed before the path was initiated.
%
The transactional interface exported by the CPC abstraction is inspired by the
work on speculative concurrency control using software transactional memory
(STM)~\cite{stm-st95}.
%, thus the  term \emph{software transactional networking}.
Our interface is however intended to model realistic network
management operations, which makes it simpler than recent
dynamic STM models~\cite{dstm}.
%In particular, the sets of rules to be installed by a policy update do not depend on the state of the
%network.

Extending the interface to dynamic policies that adapt their behavior based on
the current network state sounds like a promising research direction.  On the
other hand, our criterion of sequential composability is more complex than
traditional STM correctness properties in that it imposes restrictions not only
on the high-level interface exported to the control plane, but also on the
paths taken by the data-plane packets.
Also, we assumed that controllers are subject to failures, which is usually not
assumed by STM implementations.

@@

This paper showed that powerful synchronization primitives
from distributed computing can be implemented in Openflow. 
Our work complements existing research on the design of
distributed control planes, and also provides some missing links,
e.g., for the transactional approach taken by STN. 
We also hope that our work can nourish the ongoing discussion of
what additional in-band features may be useful for future
versions of Openflow.
